Supporting Dynamic Data Partitioning in Parallel Programs

### 2012-11-02 doc/types-pldi2013/intro.tex
updates
Edits to the intro and abstract

Machine architecture, particularly at the high
performance end of the spectrum, has recently undergone a revolution.  The
latest supercomputers are now composed of heterogeneous processors
and deep memory hierarchies.  Current programming systems for these
machines have elaborate features for describing parallelism, but 
few abstractions for describing the organization of data.  However, as the architecture
of distributed memory has become increasingly complex, the placement and movement of data has
become a first order programming concern.

@@

Previous work has explored two approaches to expressing the organization of data
in distributed memory parallel programs.
One class of work uses fully static analysis 
with no runtime overhead but limited expressivity.  Two recent examples, 
Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap into a tree of collections 
of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be fully checked statically 
(see Section~\ref{sec:related}).  The
second approach relies on fully dynamic analysis of program data that doesn't
limit expressivity but incurs runtime overhead and centralized control logic that is
difficult to scale to distributed memory machines;  examples 
include transactional memory \cite{Harris05} and thread-level speculation \cite{Steffan00}.

@@

In this work, we present the static and dynamic semantics of Legion \cite{Legion12},
a parallel programming model that efficiently supports multiple, dynamic partitions
of data. Legion's {\em logical regions} are an
abstraction capturing {\em locality} (data that will be used together,
and therefore should be colocated) and {\em independence} (disjoint data
that may be operated on in parallel).  Logical regions have 
properties that allow programmers to communicate information
about the structure of program data to the Legion implementation:
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

\item {\em privilege} and {\em coherence} properties that
describe how a logical region's data may be used by computations.
\end{itemize}

@@

Through multiple partitions of a logical region, a given datum may
belong to multiple different regions.  Thus, allowing multiple
simultaneous partitions of the same data introduces
{\em aliasing}---distinct regions may have overlapping data.\footnote{Thus
  the term {\em logical} regions: language-level regions
  name sets of data but do not imply a physical layout or placement in
  the memory hierarchy. A separate system of {\em physical regions} at
  runtime holds copies of data in the language-level logical regions
  \cite{Legion12}.}

Legion also provides {\em coherence} modes for regions, which allows
programmers to permit parallel execution even in cases where tasks in
fact do access the same data in conflicting ways.  The programmer may
specify that tasks conflicting on a region  must execute serially (in program
order, the default), {\em atomically} with respect to each other, or even
{\em simultaneously} if the programmer uses their own synchronization mechanisms
to control access (e.g., locks).

To efficiently guarantee the safety and
correctness of programs using first-class regions,
privilege and coherence properties, and arbitrary region
aliasing, Legion relies on a combination of static and
dynamic checks.  A key theme that runs through the design is that very
hard static analysis problems (specifically, alias analysis) are often both easy and
inexpensive if they can be done dynamically at the granularity of regions instead of individual
heap locations, while checks that must be done on individual values are only
cheap if done statically.  Explicating and evaluating the design of Legion's
static and dynamic semantics and its implications are the topics of this paper,
which we present as a series of contributions:
\begin{itemize}
\item We present a type system for {\em Core Legion} programs that statically
  verifies the safety of individual pointers and region
  privileges at task call boundaries (Section~\ref{subsec:coretypes}).
%  A subtle issue arises in checking privileges when regions are
%  stored in and later retrieved from heap data structures.  The type
%  system provides enough static information that the Legion runtime
%  can leverage its dynamic information about region aliasing to
%  perform dynamic privelege checks guaranteeing safety.

@@

\item We prove the soundness of Legion's static type and privelege system (Section~\ref{sec:soundness}).

>>>>

Machine architecture, particularly at the high
performance end of the spectrum, has recently undergone a revolution.  The
latest supercomputers are now composed of heterogeneous processors
and deep memory hierarchies.  Current programming systems for these
machines have elaborate features for describing parallelism, but 
few abstractions for describing the organization of data.  However, as 
distributed memory machines increase in complexity,
reasoning about the organization of program data will be imperative
for supporting the management of parallelism and the movement of data.

@@

Previous work has explored two approaches for describing the organization of data
in distributed memory parallel programs.  One class of work disallows
potential data aliasing to enable fully static analyses with no
runtime overhead, but limited expressivity. Two recent examples, 
Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap into a tree of collections 
of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be fully checked statically 
(see Section~\ref{sec:related}).  The second approach permits potential data
aliasing and uses fully dynamic analyses of program data that do not 
limit expressivity.  However, this approach incurs significant runtime overhead and requires 
centralized control logic that is difficult to scale to distributed 
memory machines; examples include transactional memory \cite{Harris05} 
and thread-level speculation \cite{Steffan00}.

@@

In this work, we present the static and dynamic semantics of Legion \cite{Legion12},
a parallel programming model that efficiently reason about the data
aliasing introduced by multiple, dynamic data partitions. 
Legion's {\em logical regions} are an abstraction capturing {\em locality} 
(data that will be used together, and therefore should be colocated) 
and {\em independence} (disjoint data that may be operated on in parallel).  
Logical regions have properties that allow programmers to communicate information
about the structure of program data to the Legion implementation:
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

\item {\em privilege} and {\em coherence} properties that
describe how a logical region's data may be used by computations (explained below).
\end{itemize}

Both dynamically computed partitions
and simultaneous partitions of the same data introduce the possibility
of {\em aliasing}---distinct regions may have overlapping data.\footnote{Thus
  the term {\em logical} regions: language-level regions
  name sets of data but do not imply a physical layout or placement in
  the memory hierarchy. A separate system of {\em physical regions} at
  runtime holds copies of data in the language-level logical regions
  \cite{Legion12}.}

@@

%Through multiple partitions of a logical region, a given datum may
%belong to multiple different regions.

To support additional parallelism in the presence of aliased data,
Legion also provides {\em coherence} modes for regions, which allows
programmers to permit relaxed models execution even in cases where tasks in
fact do access the same data.  The programmer may
specify that tasks conflicting on a region  must execute serially (in program
order, the default), {\em atomically} with respect to each other, or even
{\em simultaneously} if the programmer uses their own synchronization mechanisms
to control access (e.g., locks).

To efficiently guarantee the safety and
correctness of programs using first-class regions,
privilege and coherence properties in the presence of arbitrary region
aliasing, Legion relies on a combination of static and
dynamic checks.  A key theme that runs through the design is that 
alias analysis is a very hard static analysis problem, but is both easy and
inexpensive if they done dynamically at the granularity of regions instead of individual
heap locations, while checks that must be done on individual values are only
cheap if done statically.  Explicating and evaluating the design of Legion's
static and dynamic semantics and its implications are the topics of this paper,
which we present as a series of contributions:
\begin{itemize}
\item We present a type system for {\em Core Legion} programs that statically
  verifies the safety of individual pointers and region
  privileges at task call boundaries (Section~\ref{subsec:coretypes}).
%  A subtle issue arises in checking privileges when regions are
%  stored in and later retrieved from heap data structures.  The type
%  system provides enough static information that the Legion runtime
%  can leverage its dynamic information about region aliasing to
%  perform dynamic privilege checks guaranteeing safety.

@@

\item We prove the soundness of Legion's static type and privilege system (Section~\ref{sec:soundness}).

### 2012-11-02 doc/types-pldi2013/legion-types.tex
Edits to the intro and abstract

\begin{abstract}
Many distributed-memory parallel applications dynamically partition 
data, but current language designs either 
require fully static partitioning or rely on expensive dynamic checks.
We describe Legion, a programming system with a combination
of language features and static and dynamic checks to support 
dynamic data partitions with minimal overhead.
We describe the novel aspects of the Legion design, prove the soundness of the Legion type system,
 and show Legion type checking improves performance by up to 71\% by eliding provably
safe memory checks.  Furthermore, we show how the language design and type safety
enables a hierarchical, distributed runtime scheduling algorithm, and how
a system of data coherence annotations improves expressivity without compromising
soundness. We report results for three real-world applications running
on distributed memory machines, achieving up to 62.5X speedup on 96 GPUs 
on the Keeneland supercomputer.
%Legion is a programming model that introduces
%{\em logical regions} as a mechanism for targeting large heterogeneous machines 
%with complex memory hierarchies.  Logical regions are first-class 
%values in Legion, and programmers may {\em partition} regions in multiple different
%ways into subregions.  We present
%a type system with privileges and coherence properties for reasoning about logical regions.  
%We prove the soundness of the type system and use it to show
%that all memory accesses in Legion are safe.  Furthermore, we prove that
%type soundness enables a scalable, hierarchical
%scheduling algorithm for Legion programs.  We show that an implementation of the type checker
%for Legion programs improves performance up to 71\% by eliding
%memory checks the type system proves safe.  We also demonstrate that an implementation of the Legion 
%runtime based on the hierarchical scheduling algorithm scales well for
%three real-world applications.

>>>>

\begin{abstract}
Many distributed-memory parallel applications dynamically partition 
data in a way that makes static reasoning of data organization difficult 
due to potential data aliasing. Current language designs either 
disallow aliasing to enable static analyses of program data, 
or have permitted aliasing at the cost of expensive dynamic analyses.
We describe Legion, a programming system that relies on a hybrid system
of static and dynamic checks to support reasoning about the aliasing introduced
by dynamic data partitions with minimal overhead.  We describe the novel aspects of the Legion design, 
prove the soundness of the Legion type system, and show Legion type checking 
improves performance by up to 71\% by eliding provably safe memory checks.  
Furthermore, we show how the language design and type safety
enables a hierarchical, distributed runtime scheduling algorithm, and how
a system of data coherence annotations improves performance in the presence
of aliased data.  We report results for three real-world applications running
on distributed memory machines, achieving up to 62.5X speedup on 96 GPUs 
on the Keeneland supercomputer.
%Legion is a programming model that introduces
%{\em logical regions} as a mechanism for targeting large heterogeneous machines 
%with complex memory hierarchies.  Logical regions are first-class 
%values in Legion, and programmers may {\em partition} regions in multiple different
%ways into subregions.  We present
%a type system with privileges and coherence properties for reasoning about logical regions.  
%We prove the soundness of the type system and use it to show
%that all memory accesses in Legion are safe.  Furthermore, we prove that
%type soundness enables a scalable, hierarchical
%scheduling algorithm for Legion programs.  We show that an implementation of the type checker
%for Legion programs improves performance up to 71\% by eliding
%memory checks the type system proves safe.  We also demonstrate that an implementation of the Legion 
%runtime based on the hierarchical scheduling algorithm scales well for
%three real-world applications.
